Comparing variations in sediment transport capacity between subglacial and subaerial systems

### 2023-03-09 theory.tex
update

 B\^{a}timent G\'{e}opolis, CH-1015 Lausanne}
\title{Comparison of sediment transport capacity variations in subglacial and subaerial systems }

@@

\paragraph{Key findings:}
\begin{enumerate}
\item The sediment carrying capacity of subglacial water varies more in response to changing water discharge compared to subaerial channels, as subglacial channels are pressurized by the ice above.
\item The slowly evolving size of subglacial channels means that variations in water discharge must be accommodated by changing the velocity of water flowing through the channel.
%\marie{I can't fully understand this. Too much information is overlain here. Separate concepts and give a more linear explanation.} 
\item  As water discharge changes, differences in the sediment transport capacity response of subaerial vs subglacial channels  could substantially  impact sediment transport processes at proglacial margins as well as the interpretation of sediment transport records in glacial environments. 
\end{enumerate}

\abstract % 150 words
Sediment transport capacity depends mainly on the velocity of water flowing through a channel and the channel width over which the sediment is mobilized.
In subaerial channels, changing water discharge can be accommodated by changes in the water depth, the width of the channel that  water flows over, and water velocity.
In subglacial channels, however, water flows through a conduit with a fixed size and shape, and the water is pressurized by the overlying glacier ice.
As a result, changing water discharge may modify the water velocity, but not the  shape of the channel.
Parameterization of channel dynamics indicates that increased variability in sediment transport capacity in subglacial systems occurs across a wide range of channel geometries and flow conditions.
The increased variability in sediment transport capacity impacts sediment transport processes and the interpretation of records of sediment transport from glacierized catchments.  %\marie{This last sentence seems repetitive. I do not fully understand the results. Suggest revision after a complete update of the manuscript}

@@

\laysummary % 200 words
It is well known that sediment transport capacity responds differently to changing water discharge in subglacial and subaerial channels.
In both subglacial and subaerial channels, sediment transport capacity depends, in the first order, on 1) the channel width over-which the sediment is mobilized  and 2) the shear stress applied by flowing water on sediment at the channel bottom.
The shear stress depends largely on water velocity.
In subaerial channels, changing water discharge will change the water depth, the water velocity (controlling sediment transport capacity),  and channel width (which dictates the span over which sediment transport  will occur).
Conversely, in subglacial environments, on short times scales of hours to days, subglacial channels only experience a change in water velocity following changing water discharge, because the water flows through a subglacial conduit with a fixed geometry.
The fixed conduit geometry also means that the channel width, over-which sediment can be mobilized, does not respond to changing water discharge.
As a result, sediment transport capacity varies far more in subglacial channels compared to subaerial ones, even though changing channel width can accommodate some sediment transport variability in subaerial channels.
This increased variability has several important implications for sediment transport in glacial systems compared to subaerial ones.
%\maire{This is hard to follow. Suggest you rewrite after greater clarity in the rest of the paper.}

@@

  Glaciers expel large quantities of sediment from their termini \citep{hallet1996}.
  Changes in glacier dynamics, geomorphology, and hydrology  have prompted numerous  recent studies of  sediment transport processes in cold regions \citep[e.g.][]{zhang2022}.
  Increased sediment discharge at both the glacier and regional scale \citep{costa2017,delaney2020,li2022,vergara2022}, creates an imperative to examine the processes controlling sediment discharge from glacierized catchments.

Over long periods, processes such as glacier abrasion and quarrying sculpt landscapes and create sediment to be transported fluvially from below glaciers \citep[c.f.][]{hallet1979,iverson1993,herman2021}. 
 On shorter time periods, pressurized water transports available sediment from the glacier \citep{creyts2013,beaud2018b,delaney2019}, should enough sediment be present below the glacier (i.e. in a "transport"-limited regime).

In a "transport"-limited regime, sediment discharge responds to the sediment transport capacity or the amount of sediment that could be carried by the water. Sediment transport capacity depends on the shear stress between water and sediment it flows  over \citep{shields1936,meyer1948,engelund1967} and the width of the channel $w$ over which to mobilize sediment. The shear stress  across the channel width $\tau_t$ responds to the velocity of water $v$ flowing through at channel so that 
\begin{linenomath*}
    \begin{equation}
      \label{eq:tau_t}
        \tau_t \propto w\, v^2,
        \end{equation}
  \end{linenomath*}
  % 
  where $w$ is the width of the channel.
  Following the conservation of mass, the velocity of the water flowing through a rectangular channel is given as 
  \begin{linenomath*}
    \begin{equation}
      \label{eq:v}
  v = \frac{Q_w}{S(w)},
  \end{equation}
  \end{linenomath*}
%
where $S$ is the channel's cross-sectional area that responds to channel width $w$, and $Q_w$ is water discharge.

In subaerial channels, such as rivers with open channel flow, the depth $d$ and width $w$ of the channel evolve with changing water discharge $Q_w$ \citep{leopold1953}. During a flood with greater water discharge $Q_w$, for instance, cross-sectional area $S$ increases with channel width $w$ along with velocity $v$. As a result, the shear stress across the channel $\tau_t$ will increase in response to both width $w$ and $v$.

Yet, the response of water velocity to changing water discharge in subglacial channels is different than subaerial ones.
In subglacial systems, the water is pressurized by the ice above \citep{shreve1972}.
The size of subglacial channels responds to the creep closure of the ice above and the opening of the channel by frictional heating of water flowing through the channel \citep{rothlisberger1972}. As a result, the subglacial channel size can evolve on a longer timescale than variations in water discharge \citep[e.g.][]{iken1986,andrews2014,nanni2020}.
Changes in water discharge $Q_w$ are mainly accommodated in water velocity $v$, because the width $w$ and depth $d$ of the channel are fixed (Figure~\ref{fig:cartoon}).
It follows that greater water discharge $Q_w$ will cause the shear stress across the channel $\tau_t$ to primarily respond to the water velocity $v$, as opposed to the subaerial system where width $w$ may evolve as well.

As a result, sediment mobilization in subaerial and subglacial systems responds differently to changing water discharge.
These differences have been implicitly included in a wide range of available models that  quantify sediment transport from these environments in both subglacial and subaerial systems  \citep[e.g.][]{walder1994,tucker1997,creyts2013,wickert2019,delaney2019}.
  The divergent response of changing water discharge on sediment mobilization capacity may well impact sediment dynamics at glacier margins where flow transitions from pressurized to open channel flow \citep[e.g.][]{lane2016}.
  Additionally, the variable response of sediment transport capacity to water discharge in the two systems may impact the interpretation of records of sediment transport in glacier compared to subaerial systems \citep[e.g.][]{muller1968,richards2003}.
  Yet, the differing relationship between variations in water discharge and sediment discharge in subglacial and subaerial systems has been minimally discussed.

 This manuscript leverages parameterizations of subglacial and subaerial hydraulics to explore differences in the relationships between  water discharge, channel geometry, and water velocity. 
 With these parameterizations, the manuscript discusses the impact of water discharge variability on sediment transport capacity in idealized subglacial and subaerial systems.
  Lastly, the implications for sediment transport processes in subglacial systems, compared to subaerial ones, are discussed.

  \begin{center}
    \begin{figure}[H]
      \includegraphics[width=0.65\linewidth]{Cartoon.pdf}
      \caption{Cartoon of the situation with various parameters. Water velocity magnitude in the subglacial, $v_g$ , and subaerial ,$v_f$ , systems are shown by arrow length \marie{and lines of xxx greatest value} marks at the channel boundary. $S_g$ and $S_f$ represent the cross-sectional area in glacial and subaerial channels. Note that in the subaerial channel parameterization, rectangular channel shapes are implemented (Section~\ref{sect:fluv}).} 
      \label{fig:cartoon}
    \end{figure}
  \end{center}

@@

  \ian{check units on subaerial channel factor}
  \begin{table}[H]
    \centering
    \caption{Variables, parameters, and constants }
    \begin{tabular}{ l  c  c c }
      Name &Symbol&  Value&Units \\ \hline
      \textbf{Variables}  & & & \\
      Channel hydraulic diameter (glacier, subaerial) &  $D_{hg},\,D_{hf}$&  & $\mathrm{m}$     \\
      Width of channel floor (glacial, subaerial) & $w_{cg},w_{cf}$&  & $\mathrm{m}$     \\
      Channel cross-sectional area (glacier, subaerial) &  $S_g, S_f$& & $\mathrm{m^2}$     \\
      Water discharge (instantaneous) & $Q_w$& & $\mathrm{m^{3}\,s^{-1}}$ \\
      Representative water discharge & $Q_{w}^*$& & $\mathrm{m^{3}\,s^{-1}}$ \\
      % Gradient of hydraulic potential (glacier) &$\Psi$ & & $\mathrm{Pa\, m^{-1}}$\\
      Representative hydraulic gradient  &$\Psi^*$ & & $\mathrm{Pa\, m^{-1}}$\\
      Water velocity (glacier, subaerial)  & $v_g,\,v_{f}$& & $\mathrm{m\,s^{-1}}$ \\
      Shear stress (glacier, subaerial) & $\tau_g,\,\tau_f$&& $\mathrm{Pa \, m^{-1}}$ \\
      Width-integrated shear-stress (glacier, subaerial) & $\tau_{tg},\, \tau_{tf}$&& $\mathrm{Pa \, m^{-1}}$ \\
      Reynolds number &$R_e$& & $\mathrm{(-)}$\\
      Variable &$\Gamma$&&$\mathrm{(-)}$\\
      Variability of a variable ($\Gamma$) &$\chi$& &$\mathrm{(-)}$\\
           &&&\\

      \textbf{Parameters and Constants}  & & &\\
      Gravitational constant&$g$& $-9.81$&$\mathrm{m\,s^{-2}}$\\
      Density of water & $\rho_w$& $1000$ & $\mathrm{kg\,m^{-3}}$ \\
      Density of ice & $\rho_i$& $900$ & $\mathrm{kg\,m^{-3}}$ \\
      Kinematic viscosity of water &$\nu$& $10^{-4}$& $\mathrm{m^2\,s^{-1}}$\\
      Hooke angle of channel & $\beta$ & $\frac{\pi}{2}$ & \unit{rad}\\
      Flotation fraction & $f_f$&$0.7$& $\mathrm{(-)}$\\
      Friction factor (glacier) & $f_r$ & $4$ & $\mathrm{(-)}$ \\
      Friction factor (subaerial) & $f_p$ & $3$ & $\mathrm{(-)}$\\
      Gradient of glacier surface & $\frac{\partial z_{sg}}{\partial x}$ &$0.25$& $\mathrm{(-)}$\\
      Gradient of channel bed (subaerial) &$\frac{\partial z_c}{\partial x}$ &$0.05$& $\mathrm{(-)}$\\
      Subaerial channel factor & $k$ &$3$ & $\mathrm{m^{-2}\, s}$\\
      Channel geometry exponent &$e$& $\frac{1}{2}$&$\mathrm{(-)}$ \\
      \hline
    \end{tabular}
    \label{table:vpm}
  \end{table}

@@

The formulation of the glacial system (Section~\ref{sect:sub_mode}) requires inputs of the hydraulic gradient (Equation \ref{eq:psi}; $\frac{\partial z_{sg}}{\partial x}$, $\frac{\partial z_{c}}{\partial x}$), water discharge, $Q_w$ representative water discharge, $Q_w^*$,  Hooke angle (Equation~\ref{eq:Hf}; $\beta$), and friction factor (Equation~\ref{eq:DW}; $f_r$). \marie{Help the reader find these quantities and equations}
  Values of $\beta$ and $f_r$ are chosen so that they produce reasonable values of water velocity ($0.6-- 1.2$\,\unit{m}\,\unit{s}$^{-1}$) in accordance with measurements made through dye-tracing experiments to examine subglacial hydraulic conditions \citep[Section~\ref{sect:sub_mode}, Figure~\ref{fig:model_outs}; e.g.][]{werder2010}.

@@

  \section{Results}
  First, the subglacial hydraulics parameterization is applied to a representative water discharge ($Q_w^*$; Equation~\ref{eq:DW}) of $12$\,\unit{m}$^{3}$\,\unit{s}$^{-1}$.
  Both parameterizations are forced with a water discharge ($Q_w$) varying between $8$ and $16$ \,\unit{m}$^{3}$\,\unit{s}$^{-1}$, to represent a range of sediment transport conditions, for instance over diurnal fluctuation in water discharge. The subaerial channel width factor $k$ is $3$\,\unit{m}$^{2}$\,\unit{s}$^{-1}$, and the Hooke angle $\beta$ is $\frac{\pi}{2}$.

@@

  Both subglacial and subaerial parameterizations here assume that the distribution of velocity and shear stress of water across the channel bed is homogeneous \citep[Section~\ref{sect:sub_mode}~and~\ref{sect:fluv}][]{wainwright2015}. 
  The subaerial parameterization assumes that sediment is transported across all channel widths, as opposed to large flows such as floods when much sediment transport occurs \citep{wolman1960}.
  As a result, the variability in the subaerial system across the range of discharge values could be underestimated in the parameterization (Figure~\ref{fig:range}).

  \subsection{Geomorphic Implications}
  \label{sect:GI}
  The consistent relationship between water discharge variability and relative variability of the water velocity, shear stress, and width-integrated shear stress results from the fixed geometry in the glacier system and variable geometry in the subaerial system as they respond to changes in water discharge (Equations~\ref{eq:tau} and \ref{eq:Sf}).
  Increased variability in sediment transport capacity with respect to discharge has several implications for sediment transport in subglacial systems compared to subaerial ones.
  Shear stress is scaled to the $\frac{3}{2}$ power and relies on shear stress exceeding a threshold in sediment transport relationships  such as in \citet{meyer1948}.
  In the total sediment transport relationship \citet{engelund1967}, the shear stress is scaled to $\frac{5}{2}$ power.
  In both cases, the exponent greater than $1$ of shear stress in the sediment transport relationship magnifies the variability in sediment transport beyond the highly variable shear stress and width-integrated shear stress described above in subglacial environments (Figure~\ref{fig:model_outs}\, b and d; Figure~S1).

@@

  The response of sediment discharge caused by the greater variability in sediment transport in subglacial channels will be greatest in completely transport-limited regimes \citep[e.g.][]{kasmalkar2019}.
  In supply-limited systems, the changing sediment transport capacity (Figure~\ref{fig:model_outs}) would have a minimal effect on sediment discharge due to the absence of sediment available for transport \citep{delaney2019}.
  Yet, reaching the threshold of motion for sediment  may result in additional sediment transport and the tendency for glaciers to transport sediment in a supply-limited regime whereby sediment discharge scales with glacier sliding  \citep[e.g.][]{herman2015,koppes2015}.

  The effect of pressurized channel flow could be most pronounced during precipitation or flood events when subglacial conduits have not adapted to the water discharge flowing through them, resulting in a very large increase in sediment transport capacity and discharge \citep[e.g.][]{cowan1988,delaney2019}.
  Increased sediment discharge may result from changes to sediment access below the glacier, yet rapid increases in water velocity from water flowing through a small channel will cause a large increase in sediment transport capacity (Section~\ref{sect:sub_mode}).

  Increased variability in sediment transport may also impact sediment export from the subglacial environment into proglacial areas \citep[e.g.][]{delaney2017,perolo2018}.
  Here, velocities at the highest water discharges could generally decrease (Figure~\ref{fig:model_outs}, blue and green horizontal bars), in the transition of pressurized subglacial flow to subaerial open-channel flow as the water leaves the glacier. The decrease in water velocity potentially results in sediment deposition once sediment enters the proglacial area.
  Note that the transition from pressurized to open channel flow could occur above glacier termini when cavity closure rates are reduced and water can flow freely \citep{egli2021b}. 
  Periodic sediment deposition in these areas may even result in the transition between pressurized and subaerial flow near the glacier terminus \citep{perolo2018}.

  In large catchments and further downstream of proglacial margins, water discharge is generally less variable than at the tops of catchments, where glaciers lie \citep[c.f.][]{costa2017,vanas2017,delaney2018,hasholt2018}.
  Here, external hydrological factors further downstream of glaciers could result in smaller variations in sediment transport capacity in many subaerial systems compared to subglacial ones.
  The parameterization here shows that the smallest variability in water discharge has the largest difference in relative variability from subglacial to proglacial systems (Figure~\ref{fig:range}\, d,\,e,\,f), while the variability in the subglacial system remains higher (Figure~S1).
  Therefore, at locations such as the Greenland Ice Sheet, with smaller diurnal variations in water discharge compared to many alpine catchments \citep[c.f.][]{delaney2018,hasholt2018}, the variability in sediment transport capacity in the transition from the ice sheet to the proglacial river could be more dramatic than alpine systems as water leaves the glacier and moves downstream.
  Thus the disparity in sediment transport capacity between the subglacial and subaerial systems could be most pronounced in these catchments, potentially resulting in temporally mismatched periods between when glaciers introduce sediment to these river systems  and when the river mobilizes it (Figures~\ref{fig:range}~ and~\ref{fig:model_outs}).

@@

  Large variations in sediment transport capacity below glaciers may result in sediment mobilization and deposition processes in close spatial and temporal proximity to each other below the glacier \citep{gimbert2016,perolo2018}, in response to both evolving channel geometry and water discharge with accumulating with melt down-glacier \citep{beaud2018,delaney2019}.
  Increasing variability in sediment transport down-glacier is likely because of spatial discontinuities in sediment availability, increasing water discharge down the glacier, and the transport times of subglacial sediment \citep{williams1989,delaney2019}.
  The rapid increase and decreases in sediment transport capacity over a diurnal cycle in a subglacial system, for instance, compounds the variability already present in subaerial systems \citep{williams1989,jerolmack2010}.
  As a result, variations in sediment transport capacity could be responsible for the fluctuations in sediment discharge (``flushings'') from glaciers that are not attributed to variations in water discharge \citep[e.g.][]{richards2003,swift2021}.
  Aside from the sporadic nature of these events, variations could be the result of increased water velocity resulting from reduced channel size and therefore increased sediment transport capacity.

>>>>

  B\^{a}timent G\'{e}opolis, CH-1015 Lausanne}

\title{Comparing variations in sediment transport capacity between subglacial and subaerial systems }

@@

\paragraph{Key findings:}
\begin{enumerate}
\item The slowly evolving size of subglacial channels means that variations in water discharge must be accommodated by changing the velocity of water flowing through the channel, as opposed to increasing channel area.

\item Increased variability in water velocity results in greater variability in sediment transport capacity in subglacial systems compared to subaerial ones, although the variable width of subaerial channels reduces some of the total sediment transport capacity across the channel.

\item  As water discharge changes, differences in the sediment transport capacity response of subaerial and subglacial channels could impact sediment transport processes at proglacial margins as well as the interpretation of sediment transport records in glacial environments. 
\end{enumerate}

\abstract % 150 words
Sediment transport capacity depends mainly on the shear stress or velocity of water flowing through a channel and the channel width over which the sediment is mobilized.
In subaerial channels, changing water discharge can be accommodated by changes in the water depth, the width of the channel that  water flows over, and water velocity.
In subglacial channels, however, water flows through a conduit with a slowly evolving size, because the water is pressurized by the glacier ice.
As a result, changing water discharge may modify the water velocity, but not the  shape of the channel.
Parameterizations of channel dynamics indicate that greater variability in sediment transport capacity in subglacial systems occurs across a wide range of channel geometries and flow conditions.
The increased variability in water velocity and sediment transport capacity impacts sediment transport processes and the interpretation of records of sediment transport from glacierized catchments.

@@

\laysummary % 200 words
In both subglacial and subaerial channels, sediment transport capacity depends, in the first order, on 1) the channel width over which the sediment is mobilized  and 2) the shear stress applied by flowing water on sediment at the channel bottom, depending largely on the water velocity.
In subaerial channels, changing water discharge will change the water depth, the water velocity (controlling sediment transport capacity),  and channel width (which dictates the span over which sediment transport  will occur).
Conversely, in subglacial environments, on short times scales of hours to days, subglacial channels only experience a change in water velocity following changing water discharge, because the water flows through a subglacial conduit with a fixed geometry.
The fixed conduit geometry also means that the channel width, over-which sediment can be mobilized, does not respond to changing water discharge.
Parameterizations of the processes show that sediment transport capacity varies far more in subglacial channels compared to subaerial ones, even though changing channel width can accommodate some sediment transport variability in subaerial channels.
This increased variability in sediment transport capacity in subglacial systems, compared to subaerial ones, has several important implications for sediment transport in glacierized catchments and the interpretation of sediment discharge records.

@@

  Glaciers expel large quantities of sediment from their termini \citep{hallet1996}.
  Changes in glacier dynamics, geomorphology, and hydrology  have prompted numerous  recent studies of  sediment transport processes in cold regions \citep[e.g.][]{zhang2022}.
  Recent papers examining increased sediment discharge form glacierized regions at both the glacier and regional scale \citep{costa2017,delaney2020,li2022,vergara2022}, creates an imperative to examine the processes controlling sediment discharge from glacierized catchments.

  Over long periods, processes such as glacier abrasion and quarrying sculpt landscapes and create sediment to be transported fluvially from below glaciers \citep[c.f.][]{hallet1979,iverson2012,ugelvig2018}. 
  On shorter time periods, pressurized water transports available sediment from the glacier \citep{walder1994,creyts2013,beaud2018b}, should enough sediment be present below the glacier (i.e. in a transport-limited regime).

  In a transport-limited regime, sediment discharge responds to the sediment transport capacity or the amount of sediment that could be carried by the water. Sediment transport capacity depends on the shear stress between water and sediment it flows  over \citep{shields1936,meyer1948,engelund1967} and the width of the channel $w_c$ over which to mobilize sediment. The shear stress  across the channel width $\tau_t$ responds to the velocity of water $v$ flowing through at channel so that 
  \begin{linenomath*}
    \begin{equation}
      \label{eq:tau_t}
      \tau_t \propto w_c\, v^2,
    \end{equation}
  \end{linenomath*}
  % 
  where $w_c$ is the width of the channel.
  Following the conservation of mass, the velocity of the water flowing through a rectangular channel is given as 
  \begin{linenomath*}
    \begin{equation}
      \label{eq:v}
      v = \frac{Q_w}{S},
    \end{equation}
  \end{linenomath*}
  % 
  where $Q_w$ is water discharge,  and the denominator is the channel's cross sectional area $S$.

  In subaerial channels,with open channel flow, the cross sectional area $S$ of the channel evolves with changing water discharge $Q_w$  by changing the width of the channel along with the depth \citep{leopold1953}. During a flood with greater water discharge $Q_w$, for instance, the cross sectional area $S$ increases as the channel widens and deepens, along with water velocity $v$. As a result, the shear stress across the channel $\tau_t$ will increase in response to both the channel width  and water velocity $v$.

  Yet, the response of water velocity to changing water discharge in subglacial channels is different than subaerial ones.
  In subglacial systems, the water is pressurized by the ice above \citep{shreve1972}.
  The size of subglacial channels responds to the creep closure of the ice above and the opening of the channel by frictional heating of water flowing through the channel \citep{rothlisberger1972}.
  As a result, the subglacial channel size can evolve on a longer timescale than variations in water discharge \citep[e.g.][]{iken1986,andrews2014,nanni2020}.
  In turn, changes in water discharge $Q_w$ are mainly accommodated with water velocity $v$, because the size of the channel $S$ is fixed (Equation~\ref{eq:v} and Figure~\ref{fig:cartoon}).
  It follows that greater water discharge $Q_w$ will cause the shear stress across the channel $\tau_t$ to respond to the water velocity $v$, as opposed to the subaerial system where width evolves as well.

  As a result, sediment mobilization in subaerial and subglacial systems responds differently to changing water discharge.
  These differences have been implicitly included in a wide range of available models that  quantify sediment transport from these environments in both subglacial and subaerial systems  \citep[e.g.][]{walder1994,tucker1997,creyts2013,wickert2019,hewitt2019}.
  The divergent response of changing water discharge on sediment mobilization capacity may well impact sediment dynamics at glacier margins where flow transitions from pressurized to open channel flow \citep[e.g.][]{lane2016}.
  Additionally, the variable response of sediment transport capacity to water discharge in the two systems may impact the interpretation of records of sediment transport in glacier compared to subaerial systems \citep[e.g.][]{muller1968,richards2003,ganti2016}.
  Yet, the differing relationship between variations in water discharge and sediment discharge in subglacial and subaerial systems has been minimally discussed.

  This manuscript leverages parameterizations of subglacial and subaerial hydraulics to explore differences in the relationships between  water discharge, channel geometry, and water velocity. 
  With these parameterizations, the manuscript examines the impact of water discharge variability on sediment transport capacity in idealized subglacial and subaerial systems.
  Lastly, the implications for sediment transport processes in subglacial systems, compared to subaerial ones, are discussed.

  \begin{center}
    \begin{figure}[H]
      \includegraphics[width=0.65\linewidth]{Cartoon_text.pdf}
      \caption{Cartoon of the situation with various parameters. Water velocity magnitude in the subglacial, $v_g$ , and subaerial ,$v_f$ , systems are shown by arrow length marks at the channel boundary. $S_g$ and $S_f$ represent the cross-sectional area in subglacial and subaerial channels. The subglacial channel widths $w_{cg}$ remain steady, while the subaerial channel widths evolve $w_{cf}$. Text below panels shows different variables response to high and low water discharge ($Q_{wh}$ and $Q_{wl}$, respectively).
        Relationships between the variables presented here are given in Equations~\ref{eq:tau_t} and~\ref{eq:v}.
        Note that in the subaerial channel parameterization, rectangular channel shapes are implemented (Section~\ref{sect:fluv}).} 
      \label{fig:cartoon}
    \end{figure}
  \end{center}

@@

  \begin{table}[H]
    \centering
    \caption{Variables, parameters, and constants }
    \begin{tabular}{ l  c  c c }
      Name &Symbol&  Value&Units \\ \hline
      \textbf{Variables}  & & & \\
      Channel hydraulic diameter (glacier, subaerial) &  $D_{hg},\,D_{hf}$&  & $\mathrm{m}$     \\
      Width of channel floor (glacial, subaerial) & $w_{cg},w_{cf}$&  & $\mathrm{m}$     \\
      Channel cross-sectional area (glacier, subaerial) &  $S_g, S_f$& & $\mathrm{m^2}$     \\
      Water discharge (instantaneous) & $Q_w$& & $\mathrm{m^{3}\,s^{-1}}$ \\
      Representative water discharge & $Q_{w}^*$& & $\mathrm{m^{3}\,s^{-1}}$ \\
      Representative hydraulic gradient  &$\Psi^*$ & & $\mathrm{Pa\, m^{-1}}$\\
      Water velocity (glacier, subaerial)  & $v_g,\,v_{f}$& & $\mathrm{m\,s^{-1}}$ \\
      Shear stress (glacier, subaerial) & $\tau_g,\,\tau_f$&& $\mathrm{Pa \, m^{-1}}$ \\
      Width-integrated shear-stress (glacier, subaerial) & $\tau_{tg},\, \tau_{tf}$&& $\mathrm{Pa \, m^{-1}}$ \\
      Reynolds number &$R_e$& & $\mathrm{(-)}$\\
      Variable &$\Gamma$&&$\mathrm{(-)}$\\
      Variability of a variable ($\Gamma$) &$\chi$& &$\mathrm{(-)}$\\
           &&&\\

      \textbf{Parameters and Constants}  & & &\\
      Gravitational constant&$g$& $-9.81$&$\mathrm{m\,s^{-2}}$\\
      Density of water & $\rho_w$& $1000$ & $\mathrm{kg\,m^{-3}}$ \\
      Density of ice & $\rho_i$& $900$ & $\mathrm{kg\,m^{-3}}$ \\
      Kinematic viscosity of water &$\nu$& $10^{-4}$& $\mathrm{m^2\,s^{-1}}$\\
      Hooke angle of channel & $\beta$ & $\frac{\pi}{2}$ & \unit{rad}\\
      Flotation fraction & $f_f$&$0.7$& $\mathrm{(-)}$\\
      Friction factor (glacier) & $f_r$ & $4$ & $\mathrm{(-)}$ \\
      Friction factor (subaerial) & $f_p$ & $3$ & $\mathrm{(-)}$\\
      Gradient of glacier surface & $\frac{\partial z_{sg}}{\partial x}$ &$0.25$& $\mathrm{(-)}$\\
      Gradient of channel bed (subaerial) &$\frac{\partial z_c}{\partial x}$ &$0.05$& $\mathrm{(-)}$\\
      Subaerial channel factor & $k$ &$3$ & $\mathrm{s\,m^{-2}}$\\
      Channel geometry exponent &$e$& $\frac{1}{2}$&$\mathrm{(-)}$ \\
      \hline
    \end{tabular}
    \label{table:vpm}
  \end{table}

@@

  The formulation of the glacial system (Section~\ref{sect:sub_mode}) requires inputs of the hydraulic gradient (Equation \ref{eq:psi}; $\frac{\partial z_{sg}}{\partial x}$, $\frac{\partial z_{c}}{\partial x}$), water discharge, $Q_w$ representative water discharge, $Q_w^*$,  Hooke angle (Equation~\ref{eq:Hf}; $\beta$), and friction factor (Equation~\ref{eq:DW}; $f_r$).

@@

  \section{Results}
  First, the subglacial hydraulics parameterization is applied to a representative water discharge ($Q_w^*$; Equation~\ref{eq:DW}) of $12$\,\unit{m}$^{3}$\,\unit{s}$^{-1}$.
  Both parameterizations are forced with a water discharge ($Q_w$) varying between $8$ and $16$ \,\unit{m}$^{3}$\,\unit{s}$^{-1}$, to represent a range of sediment transport conditions, for instance over diurnal fluctuation in water discharge. The subaerial channel width factor $k$ is $3$\,\unit{m}$^{2}$\,\unit{s}$^{-1}$, and the Hooke angle $\beta$ is $\frac{\pi}{2}$.
  Values of Hooke angle $\beta$ and subglacial friction factor $f_r$ are chosen so that they produce reasonable values of water velocity ($0.6-- 1.2$\,\unit{m}\,\unit{s}$^{-1}$) in accordance with measurements made through dye-tracing experiments to examine subglacial hydraulic conditions \citep[Section~\ref{sect:sub_mode}, Figure~\ref{fig:model_outs}; e.g.][]{werder2010}.

@@

  Both subglacial and subaerial parameterizations here assume that the distribution of velocity and shear stress of water across the channel bed is homogeneous \citep[Section~\ref{sect:sub_mode}~and~\ref{sect:fluv}][]{yager2018}. 
  The subaerial parameterization assumes that sediment is transported across all channel widths, as opposed to large flows such as floods when much sediment transport occurs \citep{wolman1960}.
  As a result, the variability in the subaerial system across the range of discharge values could be underestimated in the parameterization (Figure~\ref{fig:range}).

  \subsection{Geomorphic Implications}
  \label{sect:GI}
  The relationship between water discharge variability and relative variability of the water velocity, shear stress, and width-integrated shear stress results from the fixed geometry in the glacier system and variable geometry in the subaerial system as they respond to changes in water discharge (Equations~\ref{eq:tau_t} and \ref{eq:v}).
  Increased variability in sediment transport capacity with respect to discharge has several implications for sediment transport in subglacial systems compared to subaerial ones.
  Shear stress is scaled to the $\frac{3}{2}$ power and relies on shear stress exceeding a threshold in sediment transport relationships  such as in \citet{meyer1948}.
  In the total sediment transport relationship \citet{engelund1967}, the shear stress is scaled to $\frac{5}{2}$ power.
  In both cases, the exponent greater than $1$ of shear stress in the sediment transport relationship magnifies the variability in sediment transport beyond the highly variable shear stress and width-integrated shear stress described above in subglacial environments (Figure~\ref{fig:model_outs}\, b and d; Figure~S1).

@@

  The response of sediment discharge caused by the greater variability in sediment transport in subglacial channels will be greatest in completely transport-limited regimes \citep[e.g.][]{kasmalkar2019}.
  In supply-limited systems or partially supply-limited systems, the changing sediment transport capacity (Figure~\ref{fig:model_outs}) would have a minimal effect on sediment discharge due to the absence of sediment available for transport \citep{delaney2019}.
  Yet, reaching the threshold of motion for sediment frequently  may result in additional sediment transport and the tendency for glaciers to transport sediment in a supply-limited regime whereby sediment discharge scales with glacier sliding  \citep[e.g.][]{herman2015,koppes2015}.
  Sediment exhaustion by this process may also explain the dependence of sediment discharge from the Greenland Ice Sheet on basal shear stress, a proxy for bedrock erosion, as opposed to glacier melt \citep{overeem2017}.

  The effect of pressurized channel flow could be most pronounced during precipitation or flood events when subglacial conduits have not adapted to the water discharge flowing through them, resulting in a  large increase in sediment transport capacity and discharge \citep[e.g.][]{cowan1988,delaney2019}.
  Increased sediment discharge may result from changes to sediment access below the glacier, yet rapid increases in water velocity from water flowing through a small channel will cause a large increase in sediment transport capacity (Section~\ref{sect:sub_mode}).

  Increased variability in sediment transport may also impact sediment export from the subglacial environment into proglacial areas \citep[e.g.][]{delaney2017,perolo2018}.
  Here, velocities at the highest water discharges could generally decrease (Figure~\ref{fig:model_outs}, blue and green horizontal bars), in the transition of pressurized subglacial flow to subaerial open-channel flow as the water leaves the glacier. The decrease in water velocity potentially results in sediment deposition once sediment enters the proglacial area.
  Note that the transition from pressurized to open channel flow could occur above glacier termini when cavity closure rates are reduced and water can flow freely \citep{egli2021b}. 
  Periodic sediment deposition in sediment deposition may even result in the transition between pressurized and subaerial flow near the glacier terminus \citep{perolo2018}.

  In large catchments and further downstream of proglacial margins, water discharge is generally less variable than at the tops of catchments, where glaciers lie \citep[c.f.][]{swift2005,riihimaki2005,costa2017,vanas2017,hasholt2018}.
  Here, external hydrological factors further downstream of glaciers could result in smaller variations in sediment transport capacity in many subaerial systems compared to subglacial ones.

  The parameterization here shows that the smallest variability in water discharge has the largest difference in relative variability in sediment transport parameters (Figure~\ref{fig:range}\, d,\,e,\,f).
  At locations such as the Greenland Ice Sheet, smaller diurnal variations in water discharge compared to many alpine catchments \citep[c.f.][]{swift2005,riihimaki2005,hasholt2018}.
  Therefore, the variability in sediment transport capacity in the transition from the ice sheet to the proglacial river could be more dramatic in Greenland than in alpine systems as water leaves the glacier and moves downstream.
  Thus the disparity in sediment transport capacity between the subglacial and subaerial systems could be most pronounced in these catchments, potentially resulting in temporally mismatched periods between when glaciers introduce sediment to these river systems  and when the river mobilizes it (Figures~\ref{fig:range}~ and~\ref{fig:model_outs}).

@@

  Large variations in sediment transport capacity below glaciers may result in sediment mobilization and deposition processes in close spatial and temporal proximity to each other below the glacier \citep{gimbert2016,perolo2018}.
  This sediment mobilization and deposition would occur in response to both evolving channel geometry and water discharge with accumulating with melt down-glacier \citep{beaud2018,delaney2019}.
  Increasing variability in sediment transport down-glacier is likely because of spatial discontinuities in sediment availability, increasing water discharge down the glacier, and the transport times of subglacial sediment \citep{williams1989,delaney2019}.
  The rapid increase and decreases in sediment transport capacity over a diurnal cycle in a subglacial system, for instance, compounds the variability already present in subaerial systems \citep{williams1989,jerolmack2010}.
  As a result, variations in sediment transport capacity could be responsible for the fluctuations in sediment discharge (``flushings'') from glaciers that are not attributed to variations in water discharge \citep[e.g.][]{richards2003,swift2021}.
  Aside from the sporadic nature of these events, variations could be the result of increased water velocity resulting from reduced channel size, not water discharge, therefore increasing sediment transport capacity.
